---
title: "Modelling the probability of zero sharks, one shark or more than one shark"
format: html
editor: visual
date: 2024-11-25
date-format: short
author: CBuelow
theme: minty
embed-resources: true
toc: true
toc-depth: 5
---

We start by subsetting the shark abundance data to the coral triangle to develop the model and demonstrate the approach.

##### Subset, wrangle and visualise the data

Note that covariates are selected based on Klinard et al. (in prep) DAG. Continuous covariates are mean centered and scaled, and skewed covariates are log(x+min) transformed.

```{r}
#| output: false
#| warning: false
# read in, wrangle and visualise the data
library(tidyverse)
library(GGally)
scale_2SD <- function(x) (x-mean(x, na.rm = T))/(2*sd(x, na.rm = T)) # function to mean center and scale continuous predictors (note dividing by 2 standard deviations (as recommended by Gelman))
logtrans <- function(x) log(x + (min(x[x>0], na.rm = T))) # log transform for skewed covariates
  
dat <- read.csv('data/fpdat_final.csv') |> 
  filter(region_name == "Coral Triangle") |> 
  # separate fishing restrictions into categorical variables for each limit type
  separate(col = 'fishing_restrictions', 
           into = c('limits1', 'limits2', 'limits3', 'limits4', 'limits5', 'limits6', 'limits7', 'limits8')) |> 
  mutate(gear_limits = ifelse(limits1 == 'gear' | limits2 == 'gear' | limits3 == 'gear' | limits4 == 'gear' | limits5 == 'gear' | limits8 == 'gear', 1, 0),
         species_limits = ifelse(limits1 == 'species' | limits2 == 'species' | limits3 == 'species' | limits4 == 'species' | limits5 == 'species' | limits8 == 'species', 1, 0),
         catch_limits = ifelse(limits1 == 'bag' | limits2 == 'bag' | limits3 == 'bag' | limits4 == 'bag' | limits5 == 'bag' | limits8 == 'bag', 1, 0),
         effort_limits = ifelse(limits1 == 'effort' | limits2 == 'effort' | limits3 == 'effort' | limits4 == 'effort' | limits5 == 'effort' | limits8 == 'effort', 1, 0),
         size_limits = ifelse(limits1 == 'size' | limits2 == 'size' | limits3 == 'size' | limits4 == 'size' | limits5 == 'size' | limits8 == 'size', 1, 0),
         temporal_limits = ifelse(limits1 == 'temporal' | limits2 == 'temporal' | limits3 == 'temporal' | limits4 == 'temporal' | limits5 == 'temporal' | limits8 == 'temporal', 1, 0)) |> 
  # select response and covariates of interest
  select(set_lat, set_long, set_id, reef_id, location_id, region_id, shark_maxn,
         mpa_present, mpa_age, mpa_area, mpa_compliance,
         gear_limits, species_limits, catch_limits, effort_limits, size_limits, temporal_limits,
         shark_protection_status, shark_sanctuary, HDI_2015, gov_effect_2016, population_2016, Grav_Total) |>   # make maxn an ordinal response, and the other categorical variables factors
  # put continuous covariates on same scale as binary by dividing by 2 standard deviations (as recommended by Gelman), also mean center to improve interpretation of coeffs in presence of interactions
  mutate(shark_maxn = factor(ifelse(shark_maxn > 1, 2, shark_maxn), order = T, levels = c('0','1','2')),
         mpa_compliance = ifelse(mpa_compliance == 'high', 1, 0),
         mpa_present = ifelse(mpa_present == 'yes', 1, 0),
         across(c(mpa_present:temporal_limits), ~ifelse(is.na(.), 0, .)),
         across(c(set_id:region_id, mpa_present, mpa_compliance:shark_sanctuary), factor),
         across(c(mpa_age, mpa_area, population_2016, Grav_Total), logtrans),
         across(c(mpa_age, mpa_area, HDI_2015, gov_effect_2016, population_2016, Grav_Total), scale_2SD),
         shark_protection_status = relevel(shark_protection_status, ref = "open"))
```

```{r}
head(dat) 
```

```{r}
summary(dat) 
```

**Note** that there are no shark sanctuaries in Coral triangle, so will not include as a covariate.

```{r,fig.width=9,fig.height=8, fig.align='center'}
#| error: false
#| warning: false
ggpairs(select(dat, shark_maxn:Grav_Total))
```

##### Map the data

```{r}
#| warning: false
library(sf)
library(tmap)
tmap_mode('view')

dat.sf <- dat |> 
  st_as_sf(coords = c('set_long', 'set_lat'), crs = 4326)
qtm(dat.sf, dots.col = 'shark_protection_status')
```

##### The model

We use a Bayesian multilevel, sequential ordinal regression to estimate the probability of encountering 0, 1 or \>1 shark given covariates identified in a structural causal model (Klinard et al. *in prep*).

There are different classes of ordinal models determined by the process by which ordered categories are generated in the response variable of interest (e.g., continuous vs. sequential).

Encountering 0, 1, or more than 1 shark is a sequential rather than continuous categorisation process. It is sequential because we can only encounter more than 1 shark after we have encountered 1 shark, and we can only encounter 1 shark after we have encountered 0.

Therefore, instead of using a cumulative model that assumes a single latent continuous variable for the ordered categorical response (e.g., Likert response data), we instead use a sequential model that assumes there is a latent continuous variable for every each category in the ordinal response variable (e.g., 0, 1 or \>1 sharks).

Normally, the sequential ordinal regression model would use a series of logistic regressions to model the log-odds of each category, conditional on the previous category having been met. However, we know that sequential ordinal response (0, 1 or \>1 sharks) is highly skewed towards 0 sharks. So rather than use the logit link function to model the log-odds, we will use the complementary log-log link ('cloglog') which is an alternative to logistic regression when we have rare or too frequent events and the outcome is skewed (an extreme value function). Unlike the logit, the cloglog function is asymmetrical and skewed to one side.

###### What does this mean for interpretation?

With our sequential ordinal regression we are modelling the log-odds of there being 1 or more than 1 shark encountered, conditional on there already having been 0 or 1 shark encountered, respectively. The log-odds can be converted to the probability of 0, 1, or more than one shark given the covariates included in our model.

We assumed that observing either 0 sharks, 1 shark, or more than one shark ($y$) is a sequential counting process whereby in order to observe one of the categories ($k$), one must achieve all lower categories 1 through $k$ -1. Every category $k \in {1,...,K}$ has a latent continous variable $\tilde{y_{ijkr}}$ determining the transition to the $k$ + 1th category that is $$y_{ijkr} \sim Sratio(p, u_i, a)$$

$$p_j = cloglog(\tau_j)$$ $$u_i = 0 + {\beta}X$$ $$\alpha = 1$$ $$\tau \sim N(0, 4)$$ $$\beta \sim N(0, 2)$$ $$\sigma \sim Exp(1),$$ where $y_{ijkr}$ is the ordinal response at each set ($i$; 0 sharks, 1 shark, or more than 1 shark), $p$ is a vector of $J$ probabilities ${p_1, p_2, p_3}$ corresponding to our 3 ordinal responses (0 sharks, 1 shark, more than one shark encountered). We map $J$ probabilities onto the $cloglog$ distribution (our link function) with $J - 1$ thresholds, $$\tau_1, \tau_2.$$

For identification purposes, $\beta_0$ is fixed to 0 (see third line of equation) and $alpha$ (the discrimination parameter) is set to 1.

##### Prior predictive check

First check what we can set priors for.

```{r}
#| error: false
#| warning: false
#| output: false
library(brms)
library(tidybayes)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r}
# what can we set priors for?
get_prior(shark_maxn ~  mpa_present + mpa_age + mpa_area + mpa_compliance +
             gear_limits + species_limits + catch_limits + effort_limits + size_limits + temporal_limits + HDI_2015 + gov_effect_2016 + population_2016 + shark_protection_status + Grav_Total + (1|location_id/reef_id/set_id), data = dat, family = sratio('logit'))
```

Set weakly informative priors for all parameters and do a prior predictive check.

```{r}
# set weakly informative priors and do a prior predictive check
#fit_prior <- brm(shark_maxn ~  mpa_present + mpa_age + mpa_area + mpa_compliance +
 #                  gear_limits + species_limits + catch_limits + effort_limits + size_limits + temporal_limits + HDI_2015 + gov_effect_2016 + population_2016 + shark_protection_status + Grav_Total + (1|location_id/reef_id/set_id), #+ shark_sanctuary + 
  #               prior = c(prior(normal(0, 4), class = Intercept),
   #                        prior(normal(0, 2), class = b),
    #                       prior(exponential(1), class = sd)),
     #            data = dat, family = sratio('cloglog'),
      #           iter = 4000, warmup = 1000, cores = 4, chains = 4, thin = 1,
       #          control = list(max_treedepth = 15, adapt_delta = 0.99),
        #         sample_prior = "only")

load(file = "outputs/models/ordinal-test_coral-triangle_v2.rda")
pp_check(fit_prior, type = 'bars', ndraws = 1000)
```

TODO: work on the priors, try 1 sd on the intercept and slope? Follow Soloman example but don't assume equal probability of each category

##### Fit the model with data

**Note** we can fit category-specific effects for covariates where we expect there would be different effects on the probability of encountering 0, 1, or more sharks.

```{r}
#fit <- brm(shark_maxn ~  mpa_present + mpa_age + mpa_area + mpa_compliance +
 #            gear_limits + species_limits + catch_limits + effort_limits + size_limits + temporal_limits + HDI_2015 + gov_effect_2016 + population_2016 + shark_protection_status + Grav_Total + (1|location_id/reef_id/set_id), #+ shark_sanctuary + 
  #         prior = c(prior(normal(0, 4), class = Intercept),
   #                  prior(normal(0, 2), class = b),
    #                 prior(exponential(1), class = sd)),
     #      iter = 4000, warmup = 1000, cores = 4, chains = 4, thin = 1,
      #     data = dat, family = sratio('cloglog'), 
       #    control = list(max_treedepth = 15, adapt_delta = 0.99))
# save models 
#save(fit_prior, fit, dat, file = "outputs/models/ordinal-test_coral-triangle_v2.rda")
```

**Warning messages:** 1: There were 3 chains where the estimated Bayesian Fraction of Missing Information was low. See https://mc-stan.org/misc/warnings.html#bfmi-low 2: Examine the pairs() plot to diagnose sampling problems

3: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable. Running the chains for more iterations may help. See https://mc-stan.org/misc/warnings.html#bulk-ess

Based on warnings we won't make inferences yet, but should be ok to continue developing model before final run on full data.

##### Check model convergence and fit

Get a summary of the fitted model.

```{r}
#| error: false
#| warning: false
summary(fit)
```

Visual check of the posteriors and mcmc sampling.

```{r}
#| error: false
#| warning: false
plot(fit)
```

Posterior predictive check.

```{r}
#| error: false
#| warning: false
pp_check(fit, type = "bars", ndraws = 1000)
```

Check the residuals.

```{r}
#| error: false
#| warning: false
library(patchwork)
library(tidybayes)
library(ggplot2)
theme_set(theme_tidybayes())

rq_resids <- dat |> # calculate randomised quantile residuals
  dplyr::select(-c(set_lat:set_long, region_id)) |> 
  add_predicted_draws(fit, ndraws = 100) |> # note lower number of draws - takes ages otherwise
  mutate(.prediction = ordered(levels(shark_maxn)[.prediction], levels = levels(shark_maxn))) |>
  summarise(
    p_lower = mean(.prediction < shark_maxn),
    p_upper = mean(.prediction <= shark_maxn),
    p_residual = runif(1, p_lower, p_upper),
    z_residual = qnorm(p_residual))

a <- rq_resids |> 
  ggplot(aes(x = .row, y = z_residual)) +
  geom_point()

b <- rq_resids |> 
  ggplot(aes(sample = z_residual)) +
  geom_qq() +
  geom_abline()

a + b
```

**TODO** map out the residuals to get an indication of spatial autocorrelation

##### Plot coefficients and any interesting conditional effects

Plot the coefficients.

```{r}
#| error: false
#| warning: false
mcmc_plot(fit)
```

Plot the conditional effect of MPA area (i.e., all other covariate values are held constant at their mean value).

```{r}
# note plotting 75% credible intervals here
a <- plot(conditional_effects(fit, effects = 'mpa_area', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()
b <- plot(conditional_effects(fit, effects = 'effort_limits', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()
c <- plot(conditional_effects(fit, effects = 'Grav_Total', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()
d <- plot(conditional_effects(fit, effects = 'shark_protection_status', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()

(a+b)/(c+d) 
```

##### Make counterfactual predictions

First try a management scenario where effort limits are enforced everywhere, and compare cumulative probability

```{r}
# TODO: figure out how to use tidybayes to more easily wrangle predictions and their uncertainty

# new scenario data
effort_scenario <- dat |> 
  mutate(effort_limits = 1)

# make predictions and summarise probability of each ordinal outcome at each site
base_preds <- fit |> 
  posterior_predict(ndraws = 1000, cores = 4)

colnames(base_preds) <- dat$set_id

base_preds_prob <- base_preds |> 
  data.frame() |> 
  pivot_longer(everything(), names_to = 'set_id', values_to = 'shark_maxn') |> 
  group_by(set_id, shark_maxn) |> 
  summarise(predicted_prob = n()/1000) |> 
  mutate(scenario = 'Status quo')

# number of draws (to save time - do all draws for final run)
ndraw <- 1000

effort_preds <- fit |> 
  posterior_predict(newdata = effort_scenario, ndraws = ndraw, cores = 4)
colnames(effort_preds) <- dat$set_id

effort_preds_prob <- effort_preds |> 
  data.frame() |> 
  pivot_longer(everything(), names_to = 'set_id', values_to = 'shark_maxn') |> 
  group_by(set_id, shark_maxn) |> 
  summarise(predicted_prob = n()/ndraw) |> 
  mutate(scenario = 'Effort limits enforced')

preds_prob <- bind_rows(base_preds_prob, effort_preds_prob)

# plot the cumulative probability of seeing 1 shark or more than one shark under scenarios
nsites <- length(unique(preds_prob$set_id))

preds_prob |>
  filter(shark_maxn != 1) |> 
  group_by(scenario, shark_maxn) |> 
  arrange(predicted_prob) |> 
  mutate(cumulative_probability_encounter = (cumsum(predicted_prob)/nsites)*100,
         percent_sites = ((1:n())/nsites)*100,
         shark_maxn = ifelse(shark_maxn == 2, 'Probability of 1 shark', 'Probability of > 1 shark')) |> 
  ggplot() +
  aes(x = percent_sites, y = cumulative_probability_encounter, col = factor(shark_maxn), lty = scenario) +
  geom_line() +
  ylab('Cumulative probability of encounter \n (as proportion of total)') +
  xlab('% of sites') +
  theme_classic() +
  theme(legend.title = element_blank())

```

#### Next steps

1.  Fit full model to all data, one with and without an interaction with human gravity, compare models with loo cv and WAIC. Also perhaps try making human gravity binary (e.g., high vs. low)
2.  Check for spatial autocorrelation in residuals
3.  Make counterfactual predictions under management/conservation scenarios

#### R session Info

```{r}
sessionInfo()
```
