---
title: "Modelling socio-ecological outcomes for reef sharks with conservation"
format: html
editor: visual
date: 2024-11-07
date-format: short
author: E. Lester and C. Buelow
theme: minty
---

Here we visualise the finprint data and then start developing a Bayesian hierarchical model to estimate causal effects of conservation.

#### Fishing restrictions model

We focus on building the Fishing restrictions model first. To estimate the causal effect of fishing restrictions on reef shark abundance we need to include the following reef level covariates: RES, MPA; and the following national level covariates: HDI, GOV, VOICE, SSA.

##### Wrangle and visualise the data

```{r}
#| output: false
#| warning: false
# read in, wrangle and visualise the data
library(tidyverse) 
library(GGally)

dat <- read.csv('data/fpdat_final.csv') |> 
  select(set_lat, set_long, set_id, reef_id, location_id, region_id, shark_maxn, shark_protection_status, mpa_present, HDI_2015, gov_effect_2016, voice_2016, shark_sanctuary) |> 
  mutate(mpa_present = ifelse(is.na(mpa_present), 0, 1), across(c(set_id:region_id, shark_protection_status, shark_sanctuary, mpa_present), factor)) 
```

```{r}
head(dat) 
```

```{r}
summary(dat) 
```

```{r}
#| error: false
#| warning: false
ggpairs(select(dat, shark_maxn:shark_sanctuary))
```

Note the strong +ve correlation between HDI and GOV. 
**TODO**: need to figure out what to do given using SCM to inform covariate selection.

##### Map the data

```{r}
#| warning: false
library(sf)
library(tmap)
tmap_mode('view')

dat.sf <- dat |> 
  st_as_sf(coords = c('set_long', 'set_lat'), crs = 4326)
qtm(dat.sf, dots.col = 'shark_protection_status')
```
```{r}
qtm(dat.sf, dots.col = 'mpa_present', dot.size = 3)
```

```{r}
qtm(dat.sf, dots.col = 'shark_sanctuary', dot.size = 1)
```
   