---
title: "Modelling socio-ecological outcomes for reef sharks with conservation"
format: html
editor: visual
date: 2024-11-07
date-format: short
author: E. Lester and C. Buelow
theme: minty
---

Here we visualise the finprint data and then start developing a Bayesian hierarchical model to estimate causal effects of conservation.

#### Fishing restrictions model

We focus on building the Fishing restrictions model first. To estimate the causal effect of fishing restrictions on reef shark abundance we need to include the following reef level covariates: RES, MPA; and the following national level covariates: HDI, GOV, VOICE, SSA.

##### Wrangle and visualise the data

```{r}
#| output: false
#| warning: false
# read in, wrangle and visualise the data
library(tidyverse) 
library(GGally)

dat <- read.csv('data/fpdat_final.csv') |> 
  select(set_lat, set_long, set_id, reef_id, location_id, region_id, shark_maxn, shark_protection_status, mpa_present, HDI_2015, gov_effect_2016, voice_2016, shark_sanctuary) |> 
  mutate(mpa_present = ifelse(is.na(mpa_present), 0, 1), across(c(set_id:region_id, shark_protection_status, shark_sanctuary, mpa_present), factor)) 
```

```{r}
head(dat) 
```

```{r}
summary(dat) 
```

```{r}
#| error: false
#| warning: false
ggpairs(select(dat, shark_maxn:shark_sanctuary))
```

Note the strong +ve correlation between HDI and GOV. 
**TODO**: need to figure out what to do given using SCM to inform covariate selection.

##### Map the data

```{r}
#| warning: false
library(sf)
library(tmap)
tmap_mode('view')

dat.sf <- dat |> 
  st_as_sf(coords = c('set_long', 'set_lat'), crs = 4326)
qtm(dat.sf, dots.col = 'shark_protection_status')
```
   
##### Set up the model

```{r}
#| evaluate: false
library(Hmsc)
dat.sub <- dat[1:5000,] # try on subset of data first

# model structure ------------------------------
# we will hurdle to deal with zero inflation

# set up response, fixed and random effects
Y <- as.matrix(dat.sub$shark_maxn)
XData <- dat.sub |> 
  select(shark_protection_status:shark_sanctuary)
studyDesign <- data.frame(set = droplevels(dat.sub$set_id), 
                          reef = droplevels(dat.sub$reef_id),
                          nation = droplevels(dat.sub$location_id),
                          region = droplevels(dat.sub$region_id))
rL.reef <- HmscRandomLevel(units = levels(droplevels(dat.sub$reef_id))) # note here we are creating unstructured random effects, can do a spatial one or covariate dependent ones later
rL.nation <- HmscRandomLevel(units = levels(droplevels(dat.sub$location_id)))
rL.region <- HmscRandomLevel(units = levels(droplevels(dat.sub$region_id)))

# set up mcmc sampling parameters
nchains <- 4
thin <- 2
samples <- 1000
transient <- 500*thin
verbose <- 500*thin

# loop through to set up p-a and count models and run ------------------------------

models <- list() # list for storing models
for(i in 1:3){
  if(i == 1){
    Y.pa <- 1*(Y>0)
    m <- Hmsc(Y = Y.pa, 
                        XData = XData, 
                        XFormula = ~ shark_protection_status+mpa_present+HDI_2015+gov_effect_2016+voice_2016+shark_sanctuary,
                        #TODO: get paste worknig below so can automate
                        #XFormula = paste('~', paste(colnames(XData), collapse = '+')),
                        studyDesign = studyDesign, 
                        ranLevels = list('reef' = rL.reef, 
                                         'nation' = rL.nation,
                                         'region' = rL.region),
                        distr = switch(i, "probit", "poisson", 'lognormal poisson'))
    
    models[[i]] <- sampleMcmc(m, thin = thin, samples = samples, transient = transient, nChains = nchains)
  }else {
    Y.maxn <- Y
    Y.maxn[Y.maxn==0] = NA
    m <- Hmsc(Y = Y.maxn, 
                       XData = XData, 
                        XFormula = ~ shark_protection_status+mpa_present+HDI_2015+gov_effect_2016+voice_2016+shark_sanctuary,
                        #TODO: get paste worknig below so can automate
                        #XFormula = paste('~', paste(colnames(XData), collapse = '+')),
                        studyDesign = studyDesign, 
                        ranLevels = list('reef' = rL.reef, 
                                         'nation' = rL.nation,
                                         'region' = rL.region),
                        distr = switch(i, "probit", "poisson", 'lognormal poisson'))
    
      models[[i]] = sampleMcmc(m, thin = thin, samples = samples, transient = transient, nChains = nchains, nParallel = nchains)
  }
} 
```

**TODO**

- Set up code for: 
1. checking mcmc convergence, 
2. checking structural model assumptions, 
3. evaluate model fit and predictive capacity with x-validation
4. plot up and summarise the results
