---
title: "Modelling the probability of zero sharks, one shark or more than one shark"
format: html
editor: visual
date: 2024-11-25
date-format: short
author: CBuelow
theme: minty
embed-resources: true
---

### Using a multilevel, sequential ordinal regression to model the probability of encountering 0, 1, or more than 1 shark

We start by subsetting the shark abundance data to the coral triangle to develop the model and demonstrate the approach.

##### Subset, wrangle and visualise the data

Note that covariates are selected based on Klinard et al. (in prep) DAG.

```{r}
#| output: false
#| warning: false
# read in, wrangle and visualise the data
library(tidyverse)
library(GGally)
scale_2SD <- function(x) (x-mean(x))/(2*sd(x)) # function to mean center and scale continuous predictors

dat <- read.csv('data/fpdat_final.csv') |> 
  filter(region_name == "Coral Triangle") |> 
  # separate fishing restrictions into categorical variables for each limit type
  separate(col = 'fishing_restrictions', 
           into = c('limits1', 'limits2', 'limits3', 'limits4', 'limits5', 'limits6', 'limits7', 'limits8')) |> 
  mutate(gear_limits = ifelse(limits1 == 'gear' | limits2 == 'gear' | limits3 == 'gear' | limits4 == 'gear' | limits5 == 'gear' | limits8 == 'gear', 1, 0),
         species_limits = ifelse(limits1 == 'species' | limits2 == 'species' | limits3 == 'species' | limits4 == 'species' | limits5 == 'species' | limits8 == 'species', 1, 0),
         catch_limits = ifelse(limits1 == 'bag' | limits2 == 'bag' | limits3 == 'bag' | limits4 == 'bag' | limits5 == 'bag' | limits8 == 'bag', 1, 0),
         effort_limits = ifelse(limits1 == 'effort' | limits2 == 'effort' | limits3 == 'effort' | limits4 == 'effort' | limits5 == 'effort' | limits8 == 'effort', 1, 0),
         size_limits = ifelse(limits1 == 'size' | limits2 == 'size' | limits3 == 'size' | limits4 == 'size' | limits5 == 'size' | limits8 == 'size', 1, 0),
         temporal_limits = ifelse(limits1 == 'temporal' | limits2 == 'temporal' | limits3 == 'temporal' | limits4 == 'temporal' | limits5 == 'temporal' | limits8 == 'temporal', 1, 0)) |> 
  # select response and covariates of interest
  select(set_lat, set_long, set_id, reef_id, location_id, region_id, shark_maxn,
         mpa_present, mpa_age, mpa_area, mpa_compliance,
         gear_limits, species_limits, catch_limits, effort_limits, size_limits, temporal_limits,
         shark_protection_status, shark_sanctuary, HDI_2015, gov_effect_2016, population_2016, Grav_Total) |>   # make maxn an ordinal response, and the other categorical variables factors
  # put numerical covariates on same scale as binary by dividing by 2 standard deviations (as recommended by Gelman), also mean center to improve interpretation of coeffs in presence of interactions
  mutate(shark_maxn = factor(ifelse(shark_maxn > 1, 2, shark_maxn), order = T, levels = c('0','1','2')),
         mpa_compliance = ifelse(mpa_compliance == 'high', 1, 0),
         mpa_present = ifelse(mpa_present == 'yes', 1, 0),
         across(c(mpa_present:temporal_limits), ~ifelse(is.na(.), 0, .)),
         across(c(set_id:region_id, mpa_present, mpa_compliance:shark_sanctuary), factor),
         across(c(mpa_age, mpa_area, HDI_2015, gov_effect_2016, population_2016, Grav_Total), scale_2SD))
```

```{r}
head(dat) 
```

```{r}
summary(dat) 
```
**Note** that there are no shark sanctuaries in Coral triangle, so will not include as a covariate.

```{r,fig.width=9,fig.height=8, fig.align='center'}
#| error: false
#| warning: false
ggpairs(select(dat, shark_maxn:Grav_Total))
```

##### Map the data

```{r}
#| warning: false
library(sf)
library(tmap)
tmap_mode('view')

dat.sf <- dat |> 
  st_as_sf(coords = c('set_long', 'set_lat'), crs = 4326)
qtm(dat.sf, dots.col = 'shark_protection_status')
```

##### Prior predictive check

First check what we can set priors for.
```{r}
library(brms)
library(tidybayes)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# what can we set priors for?
get_prior(shark_maxn ~  mpa_present + mpa_age + mpa_area + mpa_compliance +
             gear_limits + species_limits + catch_limits + effort_limits + size_limits + temporal_limits + HDI_2015 + gov_effect_2016 + population_2016 + shark_protection_status + Grav_Total + (1|location_id/reef_id/set_id), data = dat, family = sratio("cloglog"))
```

Set weakly informative priors for all parameters and do a prior predictive check.
```{r}
# set weakly informative priors and do a prior predictive check
#fit_prior <- brm(shark_maxn ~  mpa_present + mpa_age + mpa_area + mpa_compliance +
 #                  gear_limits + species_limits + catch_limits + effort_limits + size_limits + temporal_limits + HDI_2015 + gov_effect_2016 + population_2016 + shark_protection_status + Grav_Total + (1|location_id/reef_id/set_id), #+ shark_sanctuary + 
  #               prior = c(prior(normal(0, 4), class = Intercept),
   #                        prior(normal(0, 2), class = b),
    #                       prior(exponential(1), class = sd)),
     #            data = dat, family = sratio("cloglog"),
      #           iter = 4000, warmup = 1000, cores = 4, chains = 4, thin = 1,
       #          control = list(max_treedepth = 15, adapt_delta = 0.99),
        #         sample_prior = "only")

load(file = "outputs/models/ordinal-test_coral-triangle_v2.rda")
pp_check(fit_prior)
```

##### Fit the model with data

**Note** we can fit category-specific effects for covariates where we expect there would be different effects on the probability of encountering 0, 1, or more sharks.
```{r}
#fit <- brm(shark_maxn ~  mpa_present + mpa_age + mpa_area + mpa_compliance +
 #            gear_limits + species_limits + catch_limits + effort_limits + size_limits + temporal_limits + HDI_2015 + gov_effect_2016 + population_2016 + shark_protection_status + Grav_Total + (1|location_id/reef_id/set_id), #+ shark_sanctuary + 
  #         prior = c(prior(normal(0, 4), class = Intercept),
   #                  prior(normal(0, 2), class = b),
    #                 prior(exponential(1), class = sd)),
     #      iter = 4000, warmup = 1000, cores = 4, chains = 4, thin = 1,
      #     data = dat, family = sratio("cloglog"), 
       #    control = list(max_treedepth = 15, adapt_delta = 0.99))
# save models 
#save(fit_prior, fit, dat, file = "outputs/models/ordinal-test_coral-triangle_v2.rda")
```

**Warning messages:**
1: There were 3 chains where the estimated Bayesian Fraction of Missing Information was low. See
https://mc-stan.org/misc/warnings.html#bfmi-low 
2: Examine the pairs() plot to diagnose sampling problems
 
3: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.
Running the chains for more iterations may help. See
https://mc-stan.org/misc/warnings.html#bulk-ess 

Based on warnings we won't make inferences yet, but should be ok to continue developing model before final run on full data.

##### Check model convergence and fit

Get a summary of the fitted model.

```{r}
summary(fit)
```

Visual check of the posteriors and mcmc sampling.

```{r}
plot(fit)
```

Posterior predictive check.

```{r}
pp_check(fit)
```

Check the residuals.

```{r}
library(patchwork)
library(tidybayes)
library(ggplot2)
theme_set(theme_tidybayes())

rq_resids <- dat %>% # calculate randomised quantile residuals
  dplyr::select(-c(set_lat:set_long, region_id)) %>% 
  add_predicted_draws(fit, ndraws = 100) %>% # note lower number of draws - takes ages otherwise
  mutate(.prediction = ordered(levels(shark_maxn)[.prediction], levels = levels(shark_maxn))) %>%
  summarise(
    p_lower = mean(.prediction < shark_maxn),
    p_upper = mean(.prediction <= shark_maxn),
    p_residual = runif(1, p_lower, p_upper),
    z_residual = qnorm(p_residual))
#save(rq_resids, file = "outputs/ordinal-test_coral-triangle_v2_resids.rda")
#load("outputs/ordinal-test_coral-triangle_v2_resids.rda")

a <- rq_resids |> 
  ggplot(aes(x = .row, y = z_residual)) +
  geom_point()

b <- rq_resids |> 
  ggplot(aes(sample = z_residual)) +
  geom_qq() +
  geom_abline()

a + b
```

**TODO** map out the residuals to get an indication of spatial autocorrelation

##### Plot coefficients and any interesting conditional effects

Plot the coefficients. 

```{r}
mcmc_plot(fit)
```

Plot the conditional effect of MPA area (i.e., all other covariate values are held constant at their mean value).

```{r}
# note plotting 75% credible intervals here
a <- plot(conditional_effects(fit, effects = 'mpa_area', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()
b <- plot(conditional_effects(fit, effects = 'effort_limits', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()
c <- plot(conditional_effects(fit, effects = 'Grav_Total', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()
d <- plot(conditional_effects(fit, effects = 'shark_protection_status', categorical = T, prob = c(0.75)), plot = FALSE)[[1]] + theme_classic()

(a+b)/(c+d) 
```

##### Make counterfactual predictions

First try a management scenario where effort limits are enforced everywhere, and compare cumulative probability 

```{r}
# new scenario data
effort_scenario <- dat |> 
  mutate(effort_limits = 1)

# make predictions and summarise probability of each ordinal outcome at each site
base_preds <- fit |> 
  posterior_predict(ndraws = 1000, cores = 4)
colnames(base_preds) <- dat$set_id

base_preds_prob <- base_preds |> 
  data.frame() |> 
  pivot_longer(everything(), names_to = 'set_id', values_to = 'shark_maxn') |> 
  group_by(set_id, shark_maxn) |> 
  summarise(predicted_prob = n()/1000) |> 
  mutate(scenario = 'baseline')

effort_preds <- fit |> 
  posterior_predict(newdata = effort_scenario, ndraws = 1000, cores = 4)
colnames(effort_preds) <- dat$set_id

effort_preds_prob <- effort_preds |> 
  data.frame() |> 
  pivot_longer(everything(), names_to = 'set_id', values_to = 'shark_maxn') |> 
  group_by(set_id, shark_maxn) |> 
  summarise(predicted_prob = n()/1000) |> 
  mutate(scenario = 'effort')

preds_prob <- bind_rows(base_preds_prob, effort_preds_prob)

# plot the cumulative probability of seeing 1 shark or more than one shark under scenarios

preds_prob |> 
  filter(shark_maxn != 1) |> 
  group_by(scenario, shark_maxn) |> 
  arrange(predicted_prob) |> 
  mutate(cumulative_probability_encounter = cumsum(predicted_prob),
         number_sites = (1:n())/2560) |> 
  ggplot() +
  aes(x = number_sites, y = cumulative_probability_encounter, col = factor(shark_maxn), lty = scenario) +
  geom_line() +
  theme_classic()

```

#### Next steps

1. Fit full model to all data, one with and without an interaction with human gravity, compare models with loo cv and WAIC. Also perhaps try making human gravity binary (e.g., high vs. low)
2. Check for spatial autocorrelation in residuals
3. Make counterfactual predictions under management/conservation scenarios