---
title: "Modelling socio-ecological outcomes for reef sharks with conservation"
format: html
editor: visual
date: 2024-11-07
date-format: short
author: E. Lester and C. Buelow
theme: minty
---

Here we visualise the finprint data and then start developing a Bayesian hierarchical model to estimate causal effects of conservation.

#### Fishing restrictions model

We focus on building the Fishing restrictions model first. To estimate the causal effect of fishing restrictions on reef shark abundance we need to include the following reef level covariates: RES, MPA; and the following national level covariates: HDI, GOV, VOICE, SSA.

##### Wrangle and visualise the data

```{r}
#| output: false
#| warning: false
# read in, wrangle and visualise the data
library(tidyverse) 
library(GGally)

dat <- read.csv('data/fpdat_final.csv') |> 
  select(set_lat, set_long, set_id, reef_id, location_id, region_id, shark_maxn, shark_protection_status, mpa_present, HDI_2015, gov_effect_2016, voice_2016, shark_sanctuary) |> 
  mutate(mpa_present = ifelse(is.na(mpa_present), 0, 1), across(c(set_id:region_id, shark_protection_status, shark_sanctuary, mpa_present), factor)) 
```

```{r}
head(dat) 
```

```{r}
summary(dat) 
```

```{r}
#| error: false
#| warning: false
ggpairs(select(dat, shark_maxn:shark_sanctuary))
```

Note the strong +ve correlation between HDI and GOV. **TODO**: need to figure out what to do given using SCM to inform covariate selection.

##### Map the data

```{r}
#| warning: false
library(sf)
library(tmap)
tmap_mode('view')

dat.sf <- dat |> 
  st_as_sf(coords = c('set_long', 'set_lat'), crs = 4326)
qtm(dat.sf, dots.col = 'shark_protection_status')
```

##### Set up and fit the model

Our shark maxn response data is zero-inflated, so we will hurdle (model presence-absence of sharks, and then model maxN conditional on shark presence). We will try modelling maxN conditional on presence with a poisson distribution and a lognormal poisson in case maxN is overdispersed.

We will not fit a spatial random effect yet (it will make the models take much longer to fit), but can add later.

```{r}
#| evaluate: false
#| output: false
library(Hmsc)

# model structure ------------------------------
# we will hurdle to deal with zero inflation

# set up response, fixed and random effects
Y <- as.matrix(dat$shark_maxn)
XData <- dat |> 
  select(shark_protection_status:shark_sanctuary)
studyDesign <- data.frame(set = droplevels(dat$set_id), 
                          reef = droplevels(dat$reef_id),
                          nation = droplevels(dat$location_id),
                          region = droplevels(dat$region_id))
rL.reef <- HmscRandomLevel(units = levels(droplevels(dat$reef_id))) # note here we are creating unstructured random effects, can do a spatial one or covariate dependent ones later
rL.nation <- HmscRandomLevel(units = levels(droplevels(dat$location_id)))
rL.region <- HmscRandomLevel(units = levels(droplevels(dat$region_id)))

# set up mcmc sampling parameters
nchains <- 4
thin <- 2
samples <- 1000
transient <- 500*thin
verbose <- 500*thin

# loop through to set up p-a and count models and run ------------------------------

models <- list() # list for storing models
system.time(
for(i in 1:3){
  if(i == 1){
    Y.pa <- 1*(Y>0)
    m <- Hmsc(Y = Y.pa, 
                        XData = XData, 
                        XFormula = ~ shark_protection_status+mpa_present+HDI_2015+gov_effect_2016+voice_2016+shark_sanctuary,
                        #TODO: get paste worknig below so can automate
                        #XFormula = paste('~', paste(colnames(XData), collapse = '+')),
                        studyDesign = studyDesign, 
                        ranLevels = list('reef' = rL.reef, 
                                         'nation' = rL.nation,
                                         'region' = rL.region),
                        distr = switch(i, "probit", "poisson", 'lognormal poisson'))
    
    models[[i]] <- sampleMcmc(m, thin = thin, samples = samples, transient = transient, nChains = nchains, nParallel = nchains)
  }else {
    Y.maxn <- Y
    Y.maxn[Y.maxn==0] = NA
    m <- Hmsc(Y = Y.maxn, 
                       XData = XData, 
                        XFormula = ~ shark_protection_status+mpa_present+HDI_2015+gov_effect_2016+voice_2016+shark_sanctuary,
                        #TODO: get paste worknig below so can automate
                        #XFormula = paste('~', paste(colnames(XData), collapse = '+')),
                        studyDesign = studyDesign, 
                        ranLevels = list('reef' = rL.reef, 
                                         'nation' = rL.nation,
                                         'region' = rL.region),
                        distr = switch(i, "probit", "poisson", 'lognormal poisson'))
    
      models[[i]] = sampleMcmc(m, thin = thin, samples = samples, transient = transient, nChains = nchains, nParallel = nchains)
  }
})

# save the models
saveRDS(models, 'outputs/models/fisheries-restrictions_full_non-spatial.rds')
```

##### Check MCMC convergence

Let's check that MCMC sampling has converged so that we can trust posterior distributions of model parameters.

###### Visualise the mcmc trace plots and posterior distributions for each beta coefficient

First we will check the presence-absence model.

```{r}
library(Hmsc)
models <- readRDS('outputs/models/models-5000rows.rds')
mpost.pa <- convertToCodaObject(models[[1]]) # extract posterior distributions for beta coefficients
plot(mpost.pa$Beta)
```
###### Check effective sample size (ideally close to 4000)

```{r}
effectiveSize(mpost.pa$Beta)
```
**TODO** - effective sample sizes are a bit low, will want to up the thinning rate for final model run

###### Gelman diagnostics - potential scale reduction factors (should be close to 1)

```{r}
gelman.diag(mpost.pa$Beta)$psrf
```
Now the poisson count model (conditional on presence)

```{r}
mpost.poiss <- convertToCodaObject(models[[2]]) # extract posterior distributions for beta coefficients
plot(mpost.poiss$Beta)
```
```{r}
effectiveSize(mpost.poiss$Beta)
```
```{r}
gelman.diag(mpost.poiss$Beta)$psrf
```

Now the lognormal poisson count model (conditional on presence)

```{r}
mpost.logpoiss <- convertToCodaObject(models[[3]]) # extract posterior distributions for beta coefficients
plot(mpost.logpoiss$Beta)
```

```{r}
effectiveSize(mpost.poiss$Beta)
```

```{r}
gelman.diag(mpost.poiss$Beta)$psrf
```
So having most trouble fitting lognormal poisson. Let's check the structural model assumptions, maybe poisson is ok.


**TODO**

-   Set up code for:

1.  checking mcmc convergence,
2.  checking structural model assumptions,
3.  evaluate model fit and predictive capacity with x-validation
4.  plot up and summarise the results
