---
title: "Visualise finprint data and develop a Bayesian hierarchical modelling framework"
format: html
editor: visual
date: 2024-11-07
date-format: short
author: E. Lester and C. Buelow
theme: minty
---

### Modelling socio-ecological outcomes for reef sharks with conservation

First we will visualise the finprint data and then start developing a Bayesian hierarchical model to estimate causal effects of conservation.

#### Fishing restrictions model
We will focus on building the Fishing restrictions model first.
To estimate the causal effect of fishing restrictions on reef shark abundance we need to include the following reef level covariates: RES, MPA; and the following national level covariates: HDI, GOV, VOICE, SSA.

##### Wrangle and visualise the data
```{r}
# read in, wrangle and visualise the data
library(tidyverse) 
library(GGally) 
library(sf) 
library(Hmsc)

dat <- read.csv('data/fpdat_final.csv') |> 
  select(set_lat, set_long, set_id, reef_id, location_id, region_id, shark_maxn, shark_protection_status, mpa_present, HDI_2015, gov_effect_2016, voice_2016, shark_sanctuary) |> 
  mutate(mpa_present = ifelse(is.na(mpa_present), 0, 1), across(c(set_id:region_id, shark_protection_status, shark_sanctuary, mpa_present), factor)) 
```

```{r}
head(dat) 
```

```{r}
summary(dat) 
```

```{r}
ggpairs(select(dat, shark_maxn:shark_sanctuary))
```

Note the strong +ve correlation between HDI and GOV. 
**TODO**: need to figure out what to do given using SCM to inform covariate selection.

##### Map the data

```{r}
dat.sf \<- dat \|\> st_as_sf(coords = c('set_long', 'set_lat'), crs = 4326)
```

